---
title: "Ordinal scales and indices"
---

::: {.callout-note appearance="simple"}
This is a work in progress that is currently undergoing heavy technical editing and copy-editing
:::

## Ordinal scales

Different from the ratio (continuous data) and the nominal (discrete data) scales, the ordinal scales are structured as rank-ordered numeric classes in a finite number of classes. There are two kinds of ordinal scales: quantitative and qualitative [@Chiang2022]. There reasons for using ordinal scales include convenience and speed of rating [@madden2017].

### Quantitative ordinal scale

In the quantitative scale, each score represents a defined interval of the percentage scale. The most well-known quantitative ordinal scale is the Horsfall-Barratt (HB) scale, developed in the early 1940s when the science of plant pathology was in need of more quantitative approaches [@hebert1982]. The HB scale splits the percent scale into 12 consecutive logarithmic-based intervals of severity from 0 to 100% . The consecutive intervals increase in size from 0 to 50% and decline in size from 50 to 100%.

The divisions of the H-B scale were based on two assumptions. First, on the logarithmic relationship between the intensity of a stimulus and sensation. Second, that the rater tends to focus on objects that are of smaller size when observing objects of two colors [@madden2017]. The topic is point of controversy in the literature, as there is little experimental evidence of these assumptions. To date, evidence shows a linear relationship between visually estimated and actual severity [@nutter2006a].

To overcome the problems associated with the H-B scale [@bock2013], a new scale was recently proposed, following a linear relationship
with percentage area diseased at severities \>10% (class 6 on the
scale) [@liu2019;@chiang2014]

Let's create the HB scale as a data frame in R and show it in a table and in a graph.

```{r}
#| label: tbl-HB
#| code-fold: true
#| tbl-cap: "The Horsfal-Barrat quantitative ordinal scale used as a tool for assessing plant disease severity "
HB <- tibble::tribble(
  ~score, ~midpoint, ~'lower', ~upper,
  0,     0,  '0',  0, 
  1,   1.5,  '0+',  3, 
  2,   4.5,  '3+',  6, 
  3,   9.0,  '6+', 12,
  4,  18.5, '12+', 25,
  5,  37.5, '25+', 50,
  6,  62.5, '50+', 75,
  7,  81.5, '75+', 88,
  8,  91.0, '88+', 94,
  9,  95.5, '94+', 97,
  10, 98.5, '97+',100, 
  11,  100,  '100',100
  )
knitr::kable(HB)
```

We can visualize in a graph where the different sizes of the percent interval encompassing each score can be easily noticed.

```{r}
#| warning: false
#| message: false
#| label: fig-hb
#| fig-cap: "Ordinal scores of the Horsfal-Barrat scale"

library(ggplot2)
theme_set(theme_bw())
HB |> 
  ggplot(aes(midpoint, score))+
  geom_point(size =2)+
  geom_line()+
  scale_x_continuous(breaks = HB$upper)+
  scale_y_continuous(n.breaks =12)+
  geom_vline(aes(xintercept = 3), linetype = 2)+
  geom_vline(aes(xintercept = 6), linetype = 2)+
  geom_vline(aes(xintercept = 12), linetype = 2)+
  geom_vline(aes(xintercept = 25), linetype = 2)+
  geom_vline(aes(xintercept = 50), linetype = 2)+
  geom_vline(aes(xintercept = 75), linetype = 2)+
  geom_vline(aes(xintercept = 88), linetype = 2)+
  geom_vline(aes(xintercept = 94), linetype = 2)+
  geom_vline(aes(xintercept = 97), linetype = 2)+
  labs(x = "Percent severity", y = "HB score")
  
  
```

### Qualitative ordinal scale

In the qualitative ordinal scales, each rating provides a description of the symptoms. An example is the ordinal 0-3 scale for rating eyespot of wheat developed by [@scott1974].

| Class | Description                                                                                                 |
|------------------|------------------------------------------------------|
| 0     | uninfected                                                                                                  |
| 1     | slight eyespot (or or more small lesion occupying in total less than half of the circumference of the stem) |
| 2     | moderate eyespot (one or more lesions occupying at least half the circumference of the stem)                |
| 3     | severe eyespot (stem completely girdled by lesions; tissue softened so that lodging would really occur)     |

: Ordinal scale for rating eyespot of wheat [@scott1974]

## Disease severity index (DSI)

Usually, when quantitative or qualitative ordinal scales are used, the scores are transformed into an index on a percentage basis, such as the disease severity index (DSI). The DSI is a single number that summarizes a large amount of information on disease severity [@chester1950]. The formula for a DSI (%) can be written as follows:

$DSI = \frac{∑(class \ frequency \ ✕ \ score \  of \ rating \ class)} {total \ number \ of \ observations \ ✕ \ maximal \ disease \ index} ✕ 100$

Let's create a function named `DSI()` to automate the calculation of the disease severity index (DSI) in a series of units (e.g. leaves) for which one of the scores of the ordinal scale has been assigned during the evaluations.

```{r}
DSI <- function(unit, score, max) {
df <- data.frame(unit, score)
tab <- df |> 
janitor::tabyl(score)
tab2 <- tab |> 
  dplyr::mutate(weight = score * n) 
(sum(tab2$weight))/(nrow(df)*max)*100
}
```

The function requires three arguments:

-   unit = the vector of the number of each unit

-   score = the vector of the scores for each unit

-   max = the maximum value of the ordinal scale.

Let's use with a toy data set composed of 12 units where each received an ordinal score. The vectors were arranged as a dataframe named scores.

```{r}
unit <- c(1:12)
sample <- c(2,3,1,1,3,4,5,0,2,5,2,1)
scores <- data.frame(unit, sample)
knitr::kable(scores)

```

The ordinal score used in this example has 6 as the maximum score. The function returns the DSI value.

```{r}
DSI(scores$unit, scores$sample, 6)

```
