---
title: "Fitting gradient models"
---

::: {.callout-note appearance="simple"}
This is a work in progress that is currently undergoing heavy technical editing and copy-editing
:::

```{r message=FALSE, warning=FALSE}
#| message: false
#| warning: false
library(tidyverse)
theme_set(theme_gray(base_size = 16)) # set global theme
```

## Dataset

The hypothetical data below shows a gradient for the number of lesions counted at varying distances in meters from the source. Let's create two vectors, one for the distances $x$ and the other for the lesion count $Y$, and then a data frame by combining the two vectors.

```{r}
# create the two vectors
x <- c(0.8, 1.6, 2.4, 3.2, 4, 7.2, 12, 15.2, 21.6, 28.8)
Y <- c(184.9, 113.3, 113.3, 64.1, 25, 8, 4.3, 2.5, 1, 0.8)
grad1 <- data.frame(x, Y) # create the dataframe
grad1 # show the gradient
```

## Visualize the gradient

```{r}
#| label: fig-fit_grad1
#| fig-cap: "Hypothetical gradient of lesion count over distances from the inoculum source" 
grad1 |> 
  ggplot(aes(x, Y))+
  theme_grey(base_size = 16)+
  geom_point(size = 2)+
  geom_line()+
  labs(y = "Lesion count",
       x = "Distance (m)")
```

## Linear regression

A linear regression model is fitted to the transformed variables according to the model. The higher the coefficient of determination, the better is the fit of the model to the data.

### Exponential model

```{r}
reg_exp <- lm(log(Y) ~ x, data = grad1)
summary(reg_exp)$r.squared
```

```{r}
#| label: fig-fit_grad2
#| fig-cap: "Fit of the exponential model to the log of lesion count over distances from the inoculum source" 
grad1 |> 
  ggplot(aes(x, log(Y)))+
  theme_grey(base_size = 16)+
  geom_point(size = 2)+
  geom_line()+
  geom_abline(slope = coef(reg_exp)[[2]], intercept = coef(reg_exp)[[1]])+
 labs(y = "Log of Lesion count",
       x = "Distance (m)")

```

### Power law model 

```{r}
reg_p <- lm(log(Y) ~ log(x), data = grad1)
summary(reg_p)$r.squared
```

```{r}
#| label: fig-fit_grad3
#| fig-cap: "Fit of the power law model to the log of lesion count over log of the distance from the inoculum source" 
grad1 |> 
  ggplot(aes(log(x), log(Y)))+
  theme_grey(base_size = 16)+
  geom_point(size = 2)+
  geom_line()+
  geom_abline(slope = coef(reg_p)[[2]], intercept = coef(reg_p)[[1]])+
 labs(y = "Log of Lesion count",
       x = "Log of distance")

```

### Modified power law model

```{r}
reg_pm <- lm(log(Y) ~ log(x + 0.4), data = grad1)
summary(reg_pm)$r.squared
```


```{r}
#| label: fig-fit_grad4
#| fig-cap: "Fit of the modified power law model to the log of lesion count over log + 0.4 of the distances from the inoculum source" 
grad1 |> 
  ggplot(aes(log(x+0.4), log(Y)))+
  theme_grey(base_size = 16)+
  geom_point(size = 2)+
  geom_line()+
  geom_abline(slope = coef(reg_pm)[[2]], intercept = coef(reg_pm)[[1]])+
 labs(y = "Log of Lesion count",
       x = "Log of distance + 0.4 (m)")

```

Based on the (highest) coefficient of determination, the modified power law provided the best fit. The graphs for the fitted models confirm this conclusion.





