---
title: "Population dynamics models"
description: |
  Deterministic models of continuous time that describe epidemic progress at the population level allowing to understand, predict and compare epidemics
bibliography: references.bib
---

# Introduction

The family of models that describe the growth of epidemics, hence population dynamics model, are known as deterministic models of continuous time @chapter2017b. These models are empirically derived, meaning that they come from their fit to disease progress curve data.

By convention, time is represented by $t$ , disease intensity by $y$ and the (absolute) rate of change between two time units is represented by $\frac{dy}{dt}$ . The latter rate depends on $y$ and the apparent infection rate (sensu Vanderplank), hereafter $r$, which summarizes the effect of host, pathogen and environment on the *infection chain*.

In this tutorial, these models and their formulations are shown using R scripts to simulate the theoretical curves produced by these models. I assume that the reader is capable of opening RStudio software and reproduce the analysis by copying and pasting the codes. Hence, a very basic knowledge o R is required. Note that I use [pipes](https://r4ds.had.co.nz/pipes.html) in my programming to express a sequence of multiple operations in a more intuitive way.

Let's start loading some essential packages for programming, customizing the outputs and defining a global ggplot theme.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggthemes)
library(patchwork)
theme_set(theme_light())
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## Non-flexible models

These population dynamics models require only two parameters to predict the data to produce the curve, hence they are known as non-flexible, as opposed to the flexible ones for which there are at least one additional parameter. Let's define values for the two model parameters, which are the apparent infection rate $r$ and $y_0$, as well as parameters for the simulations. Further modifications to these values will be handled directly in the simulation (e.g. doubling infection rate, reducing initial inoculum in half, etc.).

```{r}
y0 <- 0.001 
r <- 0.06 
tmax <- 60 # maximum duration t of the epidemics
dat <- data.frame(t = seq(1:tmax), y = seq(0:1)) # define the axes
```

Now we can proceed and learn which non-flexible models are there and for which situation they are more appropriate.

### Exponential

The differential equation for the exponential model is given by

$\frac{dy}{dt} = r_E.y$,

where $r_E$ is the infection rate (subscript E for this model) and $y$ is the disease intensity. Biologically, diseased plants, or $y$, and $r_E$ at each time contribute to disease increase. The value of $\frac{dy}{dt}$ is minimal when $y = 0$ and increases exponentially with the increase in $y$. Let's check the behavior of the absolute rate over time.

```{r, echo = T}
p0 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) y0 * exp(r * t)) +
  labs(
    y = "dy/dt",
    x = "Time",
    title = "Monomolecular model",
    subtitle ="r = 0.06"
  )

p01 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) y0 * exp(r*2 * t)) +
  labs(
    y = "dy/dt",
    x = "Time",
    title = "Monomolecular model",
    subtitle ="r = 0.12"
  )
p0| p01


```

The integral for the exponential model is given by

$y = y_0 e^{r_Et}$

Let's simulate two curves by varying $r$ while fixing $y0$ and varying the latter while fixing $r_E$. We produce the two plots in *ggplot* and add the predicted curve using the \`stat_function\`.

In the left plot below, note that the infection rate was doubled ($r$ = 0.12) and in the right plot the rate was doubled for both curves but for one curve the inoculum was increase five times.

```{r echo=TRUE, message=FALSE, warning=FALSE}

p1 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) y0 * exp(r * t), linetype = 1) +
  stat_function(fun = function(t) y0 * exp(r * 2 * t), linetype = 2) +
  ylim(0, 1) +
  labs(
    title = "Exponential model",
    subtitle = "2 times r (dashed) same y0",
    x = "Time"
  )

```

```{r}
p2 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) y0 * exp(r * 2 * t), linetype = 1) +
  stat_function(fun = function(t) y0 * 5 * exp(r * 2 * t), linetype = 2) +
  ylim(0, 1) +
  labs(title = "Exponential model", x = "Time",
       subtitle = "5 times y0 (dashed) same r")

```

The two plots are then combined to produce a combo figure using the *patchwork* R pkg.

```{r}
p1 | p2
```

### Monomolecular

The differential of the monomolecular model is given by

$\frac{dy}{dt} = r_M (1-y)$

where $\frac{y}{dt}$ is the absolute rate, $r_M$ is the rate parameter of the monomolecular model and $(1-y)$ is the proportion of non-infected (healthy) individuals or host tissue. Note that $\frac{dy}{dt}$ is maximum when $y = 0$ and decreases when $y$ approaches 1. Its decline is due to decrease in the proportion of individuals or healthy sites with the increase in $y$. Any inoculum capable of infecting the host will more likely land on infected individuals or sites. Here is the curve of the absolute rate for the monomolecular model over time.

```{r echo=TRUE, fig.height=4, fig.width=6}
dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) y0 * exp(-r * t)) +
  labs(y = "dy/dt", x = "Time", title = "Monomolecular",
       subtitle = "Absolute rate over time for y0 = 0.001 and r = 0.06")
  
```

The integral of the model usually describes the temporal patterns of the monocyclic epidemics. In those, the inoculum produced during the course of the epidemics do not contribute new infections. Therefore, differentl from the exponential model, disease intensity $y$ does not affect the epidemics and so the absolute rate is proportional to $(1-y)$.

The integral of the monomolecular model is given by

$\frac{dy}{dt} = 1 - (1-y)e^{-r_Mt}$

Let's simulate two monomolecular curves with different rate parameters.

```{r}

p3 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) 1 - ((1 - y0) * exp(-r * t))) +
  stat_function(fun = function(t) 1 - ((1 - y0) * exp(-(r / 3) * t))) +
  labs(title = "Monomolecular model",
         subtitle = "Fixed y0 = 0.001", x = "Time"
       ) +
  annotate(geom = "text", x = 35, y = 0.77, label = "r = 0.06") +
  annotate(geom = "text", x = 50, y = 0.55, label = "r = 0.02")

p4 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(fun = function(t) 1 - ((1 - y0) * exp(-r / 2 * t))) +
  stat_function(fun = function(t) 1 - ((1 - (y0 * 100)) * exp(-r / 2 * t))) +
  labs(title = "Monomolecular model", 
       subtitle = "Fixed r = 0.06", x = "Time") +
  annotate(geom = "text", x = 35, y = 0.77, label = "y0 = 0.01") +
  annotate(geom = "text", x = 45, y = 0.65, label = "y0 = 0.001")

p3 | p4
```

### Logistic

The logistic model is a more elaborated version of the two previous models as it incorporates the features of them both. Its differential is given by

$\frac{dy}{dt} = r_L. y . (1 - y)$,

where $r_L$ is the infection rate of the logistic model, $y$ is the proportion of diseased individuals or host tissue and $(1-y)$ is the proportion of non-affected individuals or host area.

Biologically, $y$ in its differential equation implies that $\frac{dy}{dt}$ increases with the increase in $y$ (as in the exponential) because more disease means more inoculum. However, $(1-y)$ leads to a decrease in $\frac{dy}{dt}$ when $y$ approaches the maximum $y=1$, because the proportion of healthy individuals or host area decreases (as in the monomolecular). Therefore, $\frac{dy}{dt}$ is minimal at the onset of the epidemics, reaches a maximum when $y/2$ and declines until $y=1$.

The integral is given by

$y = \frac{1}{1 + (1-y_0).e^{-r.t}}$,

where $r_L$ is the apparent infection rate of the logistic model and $y0$ is the disease intensity at $t=0$.

```{r}
p5 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(
    linetype = 2,
    fun = function(t) 1 / (1 + ((1 - y0) / y0) * exp(-r * 2 * t))
  ) +
  stat_function(fun = function(t) 1 / (1 + ((1 - y0) / y0) * exp(-r * 4 * t))) +
  labs(title = "Logistic model", subtitle = "Fixed y0 = 0.001", x = "Time") +
  annotate(geom = "text", x = 41, y = 0.77, label = "r = 0.18") +
  annotate(geom = "text", x = 50, y = 0.10, label = "r = 0.024")


p6 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(
    linetype = 2,
    fun = function(t) 1 / (1 + ((1 - (y0 / 10)) / (y0 / 10)) * exp(-r * 3 * t))
  ) +
  stat_function(fun = function(t) 1 / (1 + ((1 - y0) / y0) * exp(-r * 3 * t))) +
  labs(title = "Logistic model", subtitle = "Fixed r = 0.24", x = "Time") +
  annotate(geom = "text", x = 35, y = 0.77, label = "y0 = 0.001") +
  annotate(geom = "text", x = 50, y = 0.10, label = "y0 = 0.0001")


p5 | p6
```

### Gompertz

The Gompertz model is similar to the logistic and provides a very good fit to several polycyclic diseases. The differential equation is given by

$\frac{dy}{dt} = r_G.[ln(1) - ln(y)]$

Differently from the logistic, the variable representing the non-infected individuals or host area is $-ln(y)$. The integral equation is given by

$y = e^{(ln(y0)).{e^{-r_G.t)}}}$,

where $r_G$ is the apparent infection rate for the Gompertz models and $y_0$ is the disease intensity at $t = 0$.

```{r}
p7 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(
    linetype = 2,
    fun = function(t) exp(log(y0) * exp(-r/2 * t))
  ) +
  stat_function(fun = function(t) exp(log(y0) * exp(-r*2 * t))) +
  labs(title = "Gompertz model", subtitle = "Fixed y0 = 0.001", x = "Time") +
  annotate(geom = "text", x = 41, y = 0.77, label = "r = 0.12") +
  annotate(geom = "text", x = 50, y = 0.10, label = "r = 0.03")


p8 <- dat %>%
  ggplot(aes(t, y)) +
  stat_function(
    linetype = 2,
    fun = function(t) exp(log(y0) * exp(-r*2 * t))
  ) +
  stat_function(fun = function(t) exp(log(y0/1000) * exp(-r*2 * t))) +
  labs(title = "Gompertz model", subtitle = "Fixed r = 0.12", x = "Time") +
  annotate(geom = "text", x = 15, y = 0.77, label = "y0 = 0.001") +
  annotate(geom = "text", x = 25, y = 0.10, label = "y0 = 0.00001")


p7 | p8

```
