---
title: "Model fitting"
---

::: {.callout-note appearance="simple"}
This is a work in progress that is currently undergoing heavy technical editing and copy-editing
:::

Now we will fit models to multiple actual disease progress curves (DPCs) data obtained from the literature again using the *epifitter* package. A few user friendly functions will help us decide which model to choose to obtain the parameters of interest and further compare the epidemics.

To illustrate, I will use two datasets available in Chapter 3 from the book, *Study of Plant Disease Epidemics* [@chapter2017b]. In the book, SAS codes are presented to perform a few analysis. We provide an alternative code for performing similar analysis, although not perfectly reproducing the results from the book.

## Non-replicated epidemics

We will compare three DPCs of the incidence of tobacco etch, a virus disease, in peppers. Evaluations of incidence were evaluated at a 7-day interval up to 49 days. The data are available in chapter 4 (page 93) [@chapter2017b]. Let's input the data manually and create a data frame. First column is the assessment time and the other columns correspond to the treatments, called groups in the book, from 1 to 3.

## Entering data

```{r}
#| warning: false
#| message: false
library(tidyverse) # essential packages 
theme_set(theme_bw(base_size = 16)) # set global theme
```

```{r}
pepper <- 
  tribble(
   ~t,  ~`1`,  ~`2`,  ~`3`,
   0,  0.08, 0.001, 0.001,
   7,  0.13,  0.01, 0.001,
  14,  0.78,  0.09,  0.01,
  21,  0.92,  0.25,  0.05,
  28,  0.99,   0.8,  0.18,
  35, 0.995,  0.98,  0.34,
  42, 0.999,  0.99,  0.48,
  49, 0.999, 0.999,  0.74
  ) 
```

## Visualize the DPCs

Before proceeding with model selection and fitting, let's visualize the three epidemics. The code below reproduces quite exactly the top plot of Fig. 4.15 (@chapter2017b page 94). The appraisal of the curves might give us a hint on which models are the best candidates.

Because the data was entered in the wide format (each DPC is in a different column) we need to reshape it to the long format. The `pivot_longer()` function will do the job of reshaping from wide to long format so we can finally use the `ggplot()` function to produce the plot.

```{r}
#| echo: true
#| label: fig-dpcs
#| fig-cap: "Disease progress curves for three tobacco etch epidemics in pepper. Reproduced from @chapter2017b page 94"
pepper |> 
  pivot_longer(2:4, names_to ="treat", values_to = "inc") |> 
  ggplot (aes(t, inc, 
              linetype = treat, 
              shape = treat, 
              group = treat))+
  geom_line(size = 1)+
  geom_point(size =3, shape = 16)+
  annotate(geom = "text", x = 15, y = 0.84, label = "1")+
  annotate(geom = "text", x = 23, y = 0.6, label = "2")+
  annotate(geom = "text", x = 32, y = 0.33, label = "3")+
  labs(y = "Disease incidence (y)",
       x = "Time (days)")+
  theme(legend.position = "none")

```

Most of the three curves show a sigmoid shape with the exception of group 3 that resembles an exponential growth, not reaching the maximum value, and thus suggesting an incomplete epidemic. We can easily eliminate the monomolecular and exponential models and decide on the other two non-flexible models: logistic or Gompertz. To do that, let's proceed to model fitting and evaluate the statistics for supporting a final decision. There are two modeling approaches for model fitting in epifitter: the **linear** or **nonlinear** parameter-estimation methods.

## Fitting: single epidemics

Among the several options offered by *epifitter* we start with the simplest one, which is fit a model to a single epidemics using the linear regression approach. For such, the `fit_lin()` requires two arguments: time (`time`) and disease intensity (`y`) each one as a vector stored or not in a dataframe.

Since we have three epidemics, `fit_lin()` will be use three times. The function produces a list object with six elements. Let's first look at the `Stats` dataframe of each of the three lists named `epi1` to `epi3`.

```{r}
library(epifitter)
epi1 <- fit_lin(time = pepper$t,  
                y = pepper$`1` )
epi1$Stats
```

```{r}
epi2 <- fit_lin(time = pepper$t,  
  y = pepper$`2` )
epi2$Stats
```

```{r}
epi3 <- fit_lin(time = pepper$t,  
  y = pepper$`3` )
epi3$Stats

```

The statistics of the model fit confirms our initial guess that the predictions by the logistic or the Gompertz are closer to the observations than predictions by the other models. There is no much difference between them based on these statistics. However, to pick one of the models, it is important to inspect the curves with the observed and predicted values to check which model is best for all curves.

## Fitting: multiple epidemics

Before looking at the prediction, let's use another handy function that allows us to simultaneously fit the models to multiple DPC data. Different from `fit_lin()`, `fit_multi()` requires the data to be structured in the long format where there is a column specifying each of the epidemics.

Let's then create a new data set called `pepper2` using the data transposing functions of the *tidyr* package.

```{r}
pepper2 <- pepper |> 
  pivot_longer(2:4, names_to ="treat", values_to = "inc")

```

Now we fit the models to all DPCs. Note that the name of the variable indicating the DPC code needs to be informed in `strata_cols` argument.

```{r}
epi_all <- fit_multi(
  time_col = "t",
  intensity_col = "inc",
  data = pepper2,
  strata_cols = "treat",
  nlin = FALSE
)
```

Now let's select the statistics of model fitting. Again, *Epifitter* ranks the models based on the CCC (the higher the better) but it is important to check the RSE as well - the lower the better. In fact, the RSE is more important when the goal is prediction.

```{r}
epi_all$Parameters |> 
  select(treat, model, best_model, RSE, CCC)
```

To be more certain about our decision, let's advance to the final step which is to produce the plots with the observed and predicted values for each assessment time by calling the `Data` dataframe of the \``epi_all` list.

```{r}
#| label: fig-fitted
#| fig-cap: "Observed (dots) and fitted (line) values for three tobacco etch epidemics in pepper"
epi_all$Data |>
 filter(model %in% c("Gompertz", "Logistic")) |> 
  ggplot(aes(time, predicted, shape = treat)) +
  geom_point(aes(time, y)) +
  geom_line() +
  facet_wrap(~ model) +
 coord_cartesian(ylim = c(0, 1)) + # set the max to 0.6
  labs(
    y = "Disease incidence",
    x = "Time (days after emergence)"
  )
```

Overall, the logistic model seems a better fit for all the curves. Let's produce a plot with the prediction error versus time.

```{r}
#| label: fig-error
#| fig-cap: "Prediction error (dotted lines) by two models fitted to the progress curves of three tobacco etch epidemics in pepper"
epi_all$Data |>
 filter(model %in% c("Gompertz", "Logistic")) |> 
  ggplot(aes(time, predicted -y, shape = treat)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0, linetype =2)+
  facet_wrap(~ model) +
 coord_cartesian(ylim = c(-0.4, 0.4)) + # set the max to 0.6
  labs(
    y = "Prediction error",
    x = "Time (days after emergence)"
  )
```

The plots above confirms the logistic model as good fit overall because the errors for all epidemics combined are more scattered around the non-error line.

```{r}
  epi_all$Parameters |>
    filter(model == "Logistic") |>
    select(treat, y0, y0_ci_lwr, y0_ci_upr, r, r_ci_lwr, r_ci_upr 
)
```

We can produce a plot for visual inference on the differences in the parameters.

```{r}
#| label: fig-params
#| fig-cap: "Estimated infection rates (left) and initial inoculum (right) by a logistic model fitted to the progress curves of three epidemics of tobacco etch on pepper"
p1 <- epi_all$Parameters |>
  filter(model == "Logistic") |>
  ggplot(aes(treat, r)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = r_ci_lwr, ymax = r_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Epidemic",
    y = "r"
  )

p2 <- epi_all$Parameters |>
  filter(model == "Logistic") |>
  ggplot(aes(treat, 1 - exp(-y0))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = y0_ci_lwr, ymax = y0_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Epidemic",
    y = "y0"
  )

library(patchwork)
p1 | p2

```

## Designed experiments

In this next section, we will work with disease data collected over time in the same plot unit (also called repeated measures) from a designed experiment for evaluating and comparing treatment effects.

Again, we will use a dataset of progress curves shown in page 98 [@chapter2017b]. The curves represent the incidence of soybean plants symptomatic for bud blight caused by tobacco streak virus. Four treatments (different planting dates) were evaluated in randomized complete block design with four replicates. There are four assessment in time for each curve. The data was stored as a csv file and will be loaded using `read_csv()` function and stored as dataframe called `budblight`.

### Loading data

```{r}
#| message: false
#| warning: false

budblight <- read_csv("https://raw.githubusercontent.com/emdelponte/epidemiology-R/main/data/bud-blight-soybean.csv")
```

Let's have a look at the first six rows of the dataset and check the data type for each column. There is an additional column representing the replicates, called block.

```{r}
budblight
```

### Visualizing the DPCs

Let's have a look at the curves and produce a combo plot figure similar to Fig. 4.17 of the book, but without the line of the predicted values.

```{r}
#| label: fig-bud1
#| fig-cap: "Disease progress curves for the incidence of budblight of soybean in Brazil for four planting dates"
#| fig-width: 8
#| fig-height: 10
p3 <- budblight |>
  ggplot(aes(
    time, y,
    group = block,
    shape = factor(block)
  )) +
  geom_point(size = 1.5) +
  ylim(0, 0.6) +
  theme(legend.position = "none")+
  facet_wrap(~treat, ncol =1)+
  labs(y = "Disease incidence",
       x = "Time (days after emergence)")

p4 <- budblight |>
  ggplot(aes(
    time, log(1 / (1 - y)),
    group = block,
    shape = factor(block)
  )) +
  geom_point(size = 2) +
  facet_wrap(~treat, ncol = 1) +
  theme(legend.position = "none")+
  labs(y = "Transformed incidence", x = "Time (days after emergence)")

p3 | p4
```

### Model fitting

Remember that the first step in model selection is the visual appraisal of the curve data linearized with the model transformation. In the case the curves represent complete epidemics (close to 100%) appraisal of the absolute rate (difference in y between two times) over time is also helpful.

For the treatments above, it looks like the curves are typical of a monocyclic disease (the case of soybean bud blight), for which the monomolecular is usually a good fit, but other models are also possible as well. For this exercise, we will use both the linear and the nonlinear estimation method.

#### Linear regression

For convenience, we use the `fit_multi()` to handle multiple epidemics. The function returns a list object where a series of statistics are provided to aid in model selection and parameter estimation. We need to provide the names of columns (arguments): assessment time (`time_col`), disease incidence (`intensity_col`), and treatment (`strata_cols`).

```{r}
lin1 <- fit_multi(
  time_col = "time",
  intensity_col = "y",
  data = budblight,
  strata_cols = "treat",
  nlin = FALSE
)
```

Let's look at how well the four models fitted the data. Epifitter suggests the best fitted model (1 to 4, where 1 is best) for each treatment. Let's have a look at the statistics of model fitting.

```{r}
lin1$Parameters |> 
select(treat, best_model, model, CCC, RSE)
```

And now we extract values for each parameter estimated from the fit of the monomolecular model.

```{r}
lin1$Parameters |>
filter(model == "Monomolecular") |>
select(treat, y0, r)

```

Now we visualize the fit of the monomolecular model (using `filter` function - see below) to the data together with the observed data and then reproduce the right plots in Fig. 4.17 from the book.

```{r}
#| label: fig-bud2
#| fig-cap: "Observed (dot) and fitted values by a monomolecular model (line) to the data on the incidence of budblight of soybean in Brazil for four planting dates"
lin1$Data |>
  filter(model == "Monomolecular") |>
  ggplot(aes(time, predicted)) +
  geom_point(aes(time, y)) +
  geom_line(size = 0.5) +
  facet_wrap(~treat) +
  coord_cartesian(ylim = c(0, 0.6)) + # set the max to 0.6
  labs(
    y = "Disease incidence",
    x = "Time (days after emergence)"
  )
```

Now we can plot the means and respective 95% confidence interval of the apparent infection rate ($r$) and initial inoculum ($y_0$) for visual inference.

```{r}
#| label: fig-bud3
#| fig-cap: "Estimates of the infection rate (left) and initial inoculum (right) from the fit of a monomolecular model  to the data on the incidence of budblight of soybean in Brazil for four planting dates"
p5 <- lin1$Parameters |>
  filter(model == "Monomolecular") |>
  ggplot(aes(treat, r)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = r_ci_lwr, ymax = r_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Epidemic",
    y = "Infection rate (r)"
  )

p6 <- lin1$Parameters |>
  filter(model == "Monomolecular") |>
  ggplot(aes(treat, 1 - exp(-y0))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = y0_ci_lwr, ymax = y0_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Time",
    y = "Initial inoculum (y0)"
  )
p5 | p6
```

#### Non-linear regression

To estimate the parameters using the non-linear approach, we repeat the same arguments in the `fit_multi` function, but include an additional argument `nlin` set to `TRUE`.

```{r}
#| message: false
#| warning: false

nlin1 <- fit_multi(
  time_col = "time",
  intensity_col = "y",
  data = budblight,
  strata_cols = "treat",
  nlin = TRUE
)
```

Let's check statistics of model fit.

```{r}
nlin1$Parameters |>
select(treat, model, CCC, RSE, best_model)
```

And now we obtain the two parameters of interest. Note that the values are not the sames as those estimated using linear regression, but they are similar and highly correlated.

```{r}
nlin1$Parameters |>
filter(model == "Monomolecular") |>
select(treat, y0, r)
```

```{r}
#| label: fig-bud4
#| fig-cap: "Estimates of the infection rate (left) and initial inoculum (right) from the fit of a monomolecular model  to the data on the incidence of budblight of soybean in Brazil for four planting dates"
p7 <- nlin1$Parameters |>
  filter(model == "Monomolecular") |>
  ggplot(aes(treat, r)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = r_ci_lwr, ymax = r_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Epidemic",
    y = "Infection rate (r)"
  )

p8 <- nlin1$Parameters |>
  filter(model == "Monomolecular") |>
  ggplot(aes(treat, y0)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = y0_ci_lwr, ymax = y0_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Epidemic",
    y = "Initial inoculum (y0)"
  )

p7 | p8
```
